---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setlocale, echo = TRUE}
Sys.setlocale(category = "LC_ALL", locale = "English")
```

## Loading and preprocessing the data
Loading data and converting to proper formats.
```{r loading, echo = TRUE}
dataTable <- read.csv("activity.csv", colClasses = "character")
dataTable$steps <- as.numeric(dataTable$steps)
dataTable$date <- as.Date(dataTable$date, "%Y-%m-%d")
dataTable$interval <- as.numeric(dataTable$interval)
head(dataTable)
```

Generate a list of unique dates from data.
```{r datelist, echo = TRUE}
dateList <- sort(dataTable[!duplicated(dataTable[,"date"]),"date"])
summary(dateList)
```
Generate a list of unique intervals from data.
```{r intervallist, echo = TRUE}
intervalList <- sort(dataTable[!duplicated(dataTable[,"interval"]),"interval"])
summary(intervalList)
```

## What is mean total number of steps taken per day?  

Histogram of steps:
```{r perday, echo = TRUE}
dtStats <- data.frame(date = character(0), steps = numeric(0))
dtStats$date <- as.Date(dtStats$date, "%Y-%m-%d")

for(currentDate in dateList)
{
  dailySteps <- dataTable[currentDate == dataTable$date, "steps"]
  dailySum <- sum(dailySteps, na.rm = TRUE)
  
  tempDT <- data.frame(date = currentDate, steps = dailySum)
  tempDT$date <- as.Date(tempDT$date, origin="1970-01-01")
  
  dtStats <- rbind(dtStats, tempDT)
}

hist(dtStats$steps, breaks = 20, main="Number of Steps", xlab="Total number of steps taken each day", ylab = "Number of Days", col="red")
```

The mean number of steps:
```{r meansteps, echo = TRUE}
mean(dtStats$steps)
```

The median number of steps:
```{r mediansteps, echo = TRUE}
median(dtStats$steps)
```

## What is the average daily activity pattern?  
```{r period, echo = TRUE}
count <- as.numeric(0)
daStats <- data.frame(interval = numeric(0), steps = numeric(0))

for(currentInterval in intervalList)
{
  count <- count + 1
  stepStore <- dataTable[currentInterval == dataTable$interval, "steps"]
  numInstanceInterval <- length(stepStore)
  avgSteps <- sum(stepStore, na.rm = TRUE) / numInstanceInterval
  
  tempDT <- data.frame(interval = currentInterval, steps = avgSteps)
  
  daStats <- rbind(daStats, tempDT)
}

plot(daStats$interval, daStats$steps,axes = F, type="l", col="green", xlab="Time", ylab="Average Number of Steps", main="Average Daily Activity Pattern")
axis(1,at=c(0,600,1200,1800,2400), label = c("0:00","6:00","12:00","18:00","24:00"))
axis(2)
```

The 5-minute interval that contains the maximum number of steps is: 
```{r maxperiod, echo = TRUE}
daStats[max(daStats$steps), "interval"]
``` 


## Imputing missing values  

Number of NA Values:
```{r nummissing, echo = TRUE}

revDataTable <- dataTable

length(revDataTable[is.na(revDataTable$steps), "steps"])
```

Strategy: Daily averages were used to fill the NA values whenever daily averages are avaliable. If daily averages are not avaliable, the averages of the 5 minute time slot across all dates for the time slot having the NA value is used.

Creating a new dataset that is equal to the original dataset but with the missing data filled in.
```{r missing, echo = TRUE}
for(currentDate in dateList)
{
  dailySteps <- revDataTable[currentDate == revDataTable$date, "steps"]
  numDailySteps <- length(dailySteps[!is.na(dailySteps)])
  if(numDailySteps > 0)
  {
    revDataTable[(currentDate == revDataTable$date) & is.na(revDataTable$steps), "steps"] <- dtStats[(currentDate == dtStats$date), "steps"] / numDailySteps
  }
}

for(currentInterval in intervalList)
{
  intervalSteps <- revDataTable[currentInterval == revDataTable$interval, "steps"]
  numIntervalSteps <- length(intervalSteps[!is.na(intervalSteps)])

  revDataTable[(currentInterval == revDataTable$interval) & is.na(revDataTable$steps), "steps"] <- daStats[(currentInterval == daStats$interval), "steps"]
}

revStats <- data.frame(date = character(0), steps = numeric(0))
revStats$date <- as.Date(revStats$date, "%Y-%m-%d")

for(currentDate in dateList)
{
  dailySteps <- revDataTable[currentDate == revDataTable$date, "steps"]
  dailySum <- sum(dailySteps, na.rm = TRUE)
  
  tempDT <- data.frame(date = currentDate, steps = dailySum)
  tempDT$date <- as.Date(tempDT$date, origin="1970-01-01")
  
  revStats <- rbind(revStats, tempDT)
}
head(revStats)
```

Histogram of the total number of steps taken each day:
```{r hismissing, echo = TRUE}
hist(revStats$steps, breaks = 20, main="Number of Steps", xlab="Total number of steps taken each day", ylab = "Number of Days", col="red")
```

The revised mean is:
```{r meanmissing, echo = TRUE}
mean(revStats$steps)
````

The revised median is: 
```{r medianmissing, echo = TRUE}
median(revStats$steps)
``` 

These values differ from the estimates from the first part of the assignment.

The impact of imputing missing data increased the mean of the total daily number of steps while the median of the total daily number of steps remained the same.

## Are there differences in activity patterns between weekdays and weekends?

Label each day as weekend or weekday.
```{r dayofweek, echo = TRUE}
numrows <- length(dateList)
dateDT <- data.frame(date = numeric(numrows), weekdays = character(numrows), daytype = character(numrows))
dateDT$date <- as.Date(dateList, origin="1970-01-01")
dateDT$weekdays <- weekdays(as.Date(dateDT$date))
dateDT$daytype <- ifelse(dateDT$weekdays %in% c("Saturday", "Sunday"),"Weekend", "Weekday")
revDataTable <- merge(revDataTable, dateDT, "date")
head(revDataTable)
```

Calcluating the average for each interval of each day type.
```{r intervalavgbyday, echo = TRUE}
avgIntervalByDay <- aggregate(revDataTable$steps, by=list(revDataTable$interval, revDataTable$daytype), mean)
colnames(avgIntervalByDay)[1] <- "interval"
colnames(avgIntervalByDay)[2] <- "daytype"
colnames(avgIntervalByDay)[3] <- "meansteps"
head(avgIntervalByDay)
tail(avgIntervalByDay)
```

Creating the plot of the average for each interval of each day type.
```{r intervalavgbydayplot, echo = TRUE}
library(lattice)
xyplot(meansteps ~ interval | daytype, data = avgIntervalByDay, layout = c(1, 2), type="l", xlab = "Interval", ylab = "Number of steps", col="green")
```